---
title: "How did storms affect the health and economy in U.S.?"
author: "Yoshiteru Kageyama"
date: "August 21, 2015"
output: html_document
---
## Synopsis
This analysis describes how storms and other severe weather events
have caused both public health and economic problems in the U.S.
communities. I got the storm data from the U.S. National Oceanic
and Atmospheric Administration's(NOAA) storm database. I also tried
tidying them up, as they contained ambiguous and incorrect data in
the several data fields. 
In this analysis, **tornado**  is the most harmful weather event
to people in the U.S.. Tornado is much more harmful than any other
weather event in the U.S. so far. **Flood**, on the other hand, 
affects its crop and properties.

## Data Processing

The data for this assignment come in the form of a CSV file which was
compressed with bzip2. I downloaded it via the RStudio environment 
and uncompressed it. I also prepared for several R packages that were
useful for the data processing and visualization.

```{r}
# Set several Internet options
setInternet2(TRUE)
local({
  r <- getOption("repos")
  r["CRAN"] <- 'http://ftp.yz.yamagata-u.ac.jp/pub/cran/'
  options(repos = r)
})

# Download the necessary R packages, if they don't exist in the system.
reqPkgs <- c("ggplot2", "dplyr", "tidyr")

for(p in reqPkgs) {
  if (!p %in% installed.packages()){
    install.packages(p)
  }
  library(p, character.only = TRUE)
}
```

```{r cache=TRUE}
# Download the storm event data.
stormDataUrl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
compStormData <- 'StormData.csv.bz2'
stormData <- 'StormData.csv'
if (!file.exists(compStormData)) {
  download.file(stormDataUrl, compStormData)
}

# Uncompress them, if the data haven't been uncompressed yet.
if (!(stormData %in% dir())) {
  print("Uncompressing data...")
  bunzip2(filename = compStormData, destname = stormData)
}
# Read the uncompressed csv data into data frame.
stormData.csv <- read.csv(file=stormData, header=TRUE)
```

### Tidying up EVTYPE data field
Since there are several strange data found in the EVTYPE field, they must be tidied up to compliant with the NOAA document.  
My strategy of cleaning the data is as follows:

* Try to correct obviously misspelled EVTYPEs (e.g., THUNDER'E'STORM...)
* Try to regard several abbreviations as original description of EVTYPE.
 (e.g., TSTM => Thunderstorm...)
* Other strange data can be ignored in this data analysis (i.e., they can remain as is).

```{r cache=TRUE}

# Try to correct obviously misspelled EVTYPEs and abbreviations.
stormData.csv$EVTYPE = toupper(stormData.csv$EVTYPE)

# list of substitution rule.
subs <- list(c("^ *ASTRONOMICAL LOW TIDE *$", "ASTRONOMICAL LOW TIDE"),
             c("^ *(AVALANCHE|AVALANCE) *$", "AVALANCHE"),
             c("^ *(BLIZZARD (SUMMARY|WEATHER)?|BLIZZARD AND .*|BLIZZARD/.*) *$", "BLIZZARD"),
             c("^ *(COASTAL|CSTL) ?FLOOD(ING)?(/EROSION)? *$", "COASTAL FLOOD"),
             c("^ *(COLD/WIND CHILL (TEMPERATURE(S)?)?|COLD/WINDS) *$", "COLD/WIND CHILL"),
             c("^ *DENSE FOG *$", "DENSE FOG"),
             c("^ *DENSE SMOKE *$", "DENSE SMOKE"),
             c("^ *DROUGHT(/EXCESSIVE HEAT)? *$", "DROUGHT"),
             c("^ *DUST (DEVEL|DEVIL|DEVIL WATERSPOUT) *$", "DUST DEVIL"),
             c("^ *DUST ?STORM(/HIGH WINDS)? *$", "DUST STORM"),
             c("^ *EXCESSIVE HEAT(/DROUGHT)? *$", "EXCESSIVE HEAT"),
             c("^ *EXTREME (COLD)?/?(WIND\ ?CHILL(S)?)?( TEMPERATURES|/BLOWING SNO)? *$", "EXTREME COLD/WIND CHILL"),
             c("^ *(FLOOD(S|ING)?|FLOODING/HEAVY RAIN|FLOOD & HEAVY RAIN|FLOOD[ /]FLASH[ /]?FLOOD(ING)?|FLOOD/FLASH|FLOOD/RAIN/WIND(S)?|FLOOD/RIVER FLOOD|FLOOD/STRONG WIND) *$", "FLOOD"),
             c("^ *FREEZING FOG *$", "FREEZING FOG"),
             c("^ *(FROST|FROST/FREEZE|FROST\\\\\\FREEZE) *$", "FROST/FREEZE"),
             c("^ *(FUNNEL(S)?|FUNNEL CLOUD[.S]?|FUNNEL CLOUD/HAIL) *$", "FUNNEL CLOUD"),
             c("^ *(HAIL|HAIL[ \\(]?[0-9.]+\\)?|HAIL (ALOFT|DAMAGE|STORM|FLOODING|FORCE WINDS)|HAIL/ICY ROADS|HAIL/WIND(S)?|HAILSTORM(S)?|HAIL\\)) *$", "HAIL"),
             c("^ *(HEAT|HEAT (DROUGHT|WAVE(S| DROUGHT)?)|HEAT/DROUGHT|HEATBURST) *$", "HEAT"),
             c("^ *((HVY|HEAVY) RAIN(S)?(/|;|AND )?(.*)?|HEAVY RAIN EFFECTS|HEAVY RAINFALL) *$", "HEAVY RAIN"),
             c("^ *HEAVY( WET| LAKE)? SNOW(( ?(AND|/|&|-)| +)[A-Z /&]*)? *$", "HEAVY SNOW"),
             c("^ *HIGH SURF( ADVISOR(Y|IES))? *$", "HIGH SURF"),
             c("^ *HIGH WIND(S)?( +([0-9]+|\\([A-Z0-9]+\\)|AND.*)|/.*)", "HIGH WIND"),
             c("^ *HURRICANE(( [A-Z]+(/[A-Z ]+))|(-[A-Z ]+)|/TYPHOON( +[A-Z]+)?)?", "HURRICANE/TYPHOON"),
             c("^ *(ICE ?STORM((/[A-Z ]+)|( AND SNOW))?|ICE/[A-Z ]+) *$","ICE STORM"),
             c("^ *LAKE(SHORE)? FLOOD *$","LAKESHORE FLOOD"),
             c("^ *LAKE-?EFFECT SNOW *$","LAKE-EFFECT SNOW"),
             c("^ *(LIGHTING|LIGHTNING|LIGNTNING)([A-Z ./]+)?", "LIGHTNING"),
             c("^ *MARINE HAIL *$", "MARINE HAIL"),
             c("^ *MARINE HIGH WIND *$", "MARINE HIGH WIND"),
             c("^ *MARINE STRONG WIND *$", "MARINE STRONG WIND"),
             c("^ *MARINE (THUNDERSTORM|TSTM) WIND *$", "MARINE THUNDERSTORM WIND"),
             c("^ *RIP CURRENT(S([ /]HEAVY SURF)?)? *$", "RIP CURRENT"),
             c("^ *SEICHE *$", "SEICHE"),
             c("^ *(SLEET(/[A-Z /]+)?|SLEET (& )?[A-Z ]+) *$", "SLEET"),
             c("^ *STORM SURGE(/TIDE)? *$", "STORM SURGE/TIDE"),
             c("^ *STRONG WIND(S)?( GUST)? *$", "STRONG WIND"),
             c("^ *(THUNDERESTORM|THUNDERSTORM|THUNDERSTROM|THUNDERTORM|THUNDERSORM)(S)?(\\ +(WIND|W IND)(S)?([/ ]+[A-Z ]+)?)? *$", "THUNDERSTORM WINDS"),
             c("^ *(THUNERSTORM|THUNDERSTORM|THUNDESTORM|THUNDERTSORM)\\ *WIND(S)?\\ *(\\.|\\(?G[0-9]+\\)?|[0-9]+( ?MPH(\\.)?)?)? *$", "THUNDERSTORM WINDS"),
             c("^ *THUNDERSTORM(W)?(\\ ([0-9]+|WIND(S|SS)?|WINS))?", "THUNDERSTORM WINDS"),
             c("^ *THUNDERSTORM WINDS[A-Z ]+ *$", "THUNDERSTORM WINDS"),
             c("^ *TSTM\\ WIND\\ +\\(?G?[0-9]+\\)? *$", "THUNDERSTORM WINDS"),
             c("^ *TSTM (WIND(S)?|WND)(/[A-Z]+| [A-Z ]+)? *$", "THUNDERSTORM WINDS"),
             c("^ *TSTM(W) *$", "THUNDERSTORM WINDS"),
             c("^ *(TORNADO(S|ES)?|TORNAO|TORNDAO) *$", "TORNADO"),
             c("^ *TORNADO(\\ [A-Z0-9]+|/[A-Z]+|ES,[A-Z, ]+) *$", "TORNADO"),
             c("^ *TROPICAL DEPRESSION *$", "TROPICAL DEPRESSION"),
             c("^ *TROPICAL STORM(\\ +[A-Z]+)? *$", "TROPICAL STORM"),
             c("^ *TSUNAMI *$", "TSUNAMI"),
             c("^ *VOLCANIC ASH(\\ ?[A-Z]+)? *$", "VOLCANIC ASH"),
             c("^ *(WATER|WAYTER) ?SPOUT(S)?((-[A-Z]*)?|/[A-Z ]*|\\ [A-Z ]+)? *$","WATERSPOUT"),
             c("^ *(WILD\\ ?FIRE(S)|WILD/FOREST FIRE(S)?) *$", "WILDFIRE"),
             c("^ *WINTER STORM(S)?([/ ]?[A-Z ]+)? *$","WINTER STORM"),
             c("^ *WINTER WEATHER([/ ]MIX)? *$","WINTER WEATHER")
             )

# Substitute matched data in EVTYPE data field
for (s in subs) {
    stormData.csv$EVTYPE = gsub(pattern = s[[1]], 
                                replacement = s[[2]],
                                x = stormData.csv$EVTYPE)
}
```

### Calculate PROPDMG and CROPDMG data fields according to CROPDMGEXP/PROPDMGEXP.
Since there are CROPDMGEXP and PROPDMGEXP data fields in the storm data.
We must calculate PROPDMG and CROPDMG data according to PROPDMGEXP and CROPDMGEXP
data.
```{r cache=TRUE}
library(dplyr)

# Tidying up the other data fields.
stormData.csv <- stormData.csv %>%
    mutate(INJURIES   = as.numeric(INJURIES),
           FATALITIES = as.numeric(FATALITIES),
           CROPDMG    = as.numeric(CROPDMG),
           PROPDMG    = as.numeric(PROPDMG),
           CROPDMG = 
               ifelse(CROPDMGEXP == "K" | CROPDMGEXP == "k", CROPDMG * 1000,
                      ifelse(CROPDMGEXP == "M" | CROPDMGEXP == "m", CROPDMG * 1000000,
                             ifelse(CROPDMGEXP == "B" | CROPDMGEXP == "b", CROPDMG * 1000000000, 
                                    CROPDMG))),
           PROPDMG =                
               ifelse(PROPDMGEXP == "K" | PROPDMGEXP == "k", PROPDMG * 1000,
                      ifelse(PROPDMGEXP == "M" | PROPDMGEXP == "m", PROPDMG * 1000000,
                             ifelse(PROPDMGEXP == "B" | PROPDMGEXP == "b", PROPDMG * 1000000000, 
                                    PROPDMG)))
    )
```

## Results
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r cache=TRUE}
library(ggplot2)
library(tidyr)

# acquire only the data necessary to show how weather events affect human health,
# and aggregated them for each EVTYPE.
stormData.inj = select(stormData.csv, INJURIES, FATALITIES, EVTYPE)
stormData.inj = aggregate(. ~ EVTYPE, stormData.inj, sum)
stormData.inj <- mutate(stormData.inj, TOTAL_DAMAGED = INJURIES + FATALITIES) %>%
    filter(TOTAL_DAMAGED > 1000) %>%
    arrange(desc(TOTAL_DAMAGED)) %>%
    gather(DAMAGE_TYPE, NUM_INJ, INJURIES:FATALITIES)

head(stormData.inj, n = 10)

# plot the stormData.
gp = ggplot(stormData.inj, aes(x = reorder(EVTYPE, -TOTAL_DAMAGED),
                               y = NUM_INJ, fill = DAMAGE_TYPE)) + 
    geom_bar(stat = "identity") + 
    xlab("Storm Event Type") +
    ylab("Number of damaged people") +
    ggtitle("Total number of damaged people by storm events") +
    theme(axis.text.x=element_text(angle=45,hjust=1))
plot(gp)

```

In this figure, **tornado, thunderstorm winds, excessive heat, flood, and lightning**
are the top-5 most harmful weather events to people in the U.S..

2. Across the United States, which types of events have the greatest economic consequences?

```{r cache=TRUE}
# acquire only the data necessary to show how weather events affect crop and
# properties in the U.S.
stormData.propdmg = select(stormData.csv, PROPDMG, CROPDMG, EVTYPE)
stormData.propdmg = aggregate(. ~ EVTYPE, stormData.propdmg, function(x) {
    total <- sum(x)
    return(total / 1000000000)
})
stormData.propdmg <- mutate(stormData.propdmg, TOTAL_DAMAGED = PROPDMG + CROPDMG) %>%
    filter(TOTAL_DAMAGED > 10) %>%
    arrange(desc(TOTAL_DAMAGED)) %>%
    gather(DAMAGE_TYPE, AMT_DMG, PROPDMG:CROPDMG)

head(stormData.propdmg, n = 10)

# plot the storm data.
gp = ggplot(stormData.propdmg, aes(x = reorder(EVTYPE, -TOTAL_DAMAGED),
                               y = AMT_DMG, fill = DAMAGE_TYPE)) + 
    geom_bar(stat = "identity") + 
    xlab("Storm Event Type") +
    ylab("Total Amount of Damage (Billion dollars)") +
    ggtitle("Total Amount of Crop/Property Damage") +
    theme(axis.text.x=element_text(angle=45,hjust=1))
plot(gp)

```

In this figure, **flood** affects the crop and properties most in the U.S.
